---
title: "Impact of PM2.5 Exposure on Low Birth Weight in Ulaanbaatar from 2016–2025"
author: "Ahmadi, Barua, Bhuyan, Karayel"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
library(lubridate)
library(readr)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(kableExtra)
library(scales)
library(broom)
library(ggpubr)
library(here)
library(janitor)
library(forecast)
library(car)
```
**Rationale and Research Questions**
Air pollution, particularly PM2.5, remains a significant public health concern in Ulaanbaatar.  
This project examines whether exposure to elevated PM2.5 concentrations during pregnancy is associated with an increased risk of low birth weight (LBW) among newborns between 2016 and 2025.

**Research Questions:**
- What are the seasonal and long-term trends in PM2.5 concentrations in Ulaanbaatar?
- Is there a measurable association between PM2.5 exposure levels and low birth weight rates?
- Does the timing of PM2.5 exposure during pregnancy (examined through lagged exposure models) influence birth outcomes?

---

**Dataset Information**

**Data Sources**
#Birth data consisted of monthly counts of live births and low birth weight (LBW) births for each district in Ulaanbaatar. Air pollution data included daily and monthly averages of PM2.5 concentrations, aggregated to monthly values. These datasets were merged by month and district for analysis.


## Data Cleaning and Wrangling

- **PM2.5 data**:
  - Combined annual CSV files across multiple years.
  - Replaced invalid `-999` entries with `NA`.
  - Aggregated hourly readings into daily and monthly averages to align with birth record reporting.

- **Birth outcomes data**:
  - Reshaped from wide format to long format for time series analysis.
  - Merged live births and low birth weight counts into a unified dataset.

- **Final merged dataset**:
  - Monthly PM2.5 exposure data was linked with corresponding monthly birth outcome data by date.

## Dataset Structure

| Dataset           | Key Variables                         | Notes                            |
|-------------------|----------------------------------------|----------------------------------|
| PM2.5 Pollution    | DateTime, Raw Concentration, AQI       | Aggregated to monthly averages   |
| Birth Outcomes     | Date, Aimag, Low Birth Weight count, Live Births | Cleaned and merged records |


---


```{r Importing the data}

#No we load and prepare the raw PM2.5 and birth outcome datasets as described previously.

# Define file paths for PM2.5 data
years <- 2015:2025
pm25_files <- paste0("Data/Raw/Ulaanbaatar_PM2.5_", years, "_YTD.csv")
names(pm25_files) <- years

# Read and bind all PM2.5 files
pm25_all <- map_dfr(pm25_files, ~ read_csv(.x, show_col_types = FALSE))

# Clean and convert -999 to NA
pm25_all <- pm25_all %>%
  mutate(across(where(is.numeric), ~ na_if(., -999))) %>%
  clean_names() %>%
  rename(DateTime = date_lt) %>%
  mutate(
    DateTime = parse_date_time(DateTime, orders = "ymd IMp"),
    Date     = as_date(DateTime)
  )

# Load birth outcome data
birth_weight_low <- read_csv("Data/Raw/BIRTH WEIGTH LOWER THAN 2500 GRAMS.csv")
live_births      <- read_csv("Data/Raw/LIVE BIRTHS.csv")

# Clean numeric columns in live births (in case of commas)
for (col in names(live_births)[-1]) {
  live_births[[col]] <- as.numeric(gsub(",", "", live_births[[col]]))
}

# Reshape both datasets to long format
birth_weight_long <- birth_weight_low %>%
  pivot_longer(-Aimag, names_to = "Month", values_to = "Low_Birth_Weight") %>%
  mutate(Month = gsub("^X", "", Month))

live_births_long <- live_births %>%
  pivot_longer(-Aimag, names_to = "Month", values_to = "Live_Births") %>%
  mutate(Month = gsub("^X", "", Month))

# Merge and finalize
births_merged <- left_join(birth_weight_long, live_births_long, by = c("Aimag", "Month")) %>%
  mutate(Date = ym(Month)) %>%
  select(Aimag, Date, Low_Birth_Weight, Live_Births)

```
**Exploratory Analysis and Visualizations** 

```{r, Daily and Monthly Aggregation of PM2.5}

#We start by analyzing the structure, quality, and time-related aspects of the air pollution data.


# Daily average
pm25_daily <- pm25_all %>%
  group_by(Date) %>%
  summarize(
    raw_conc_daily = mean(raw_conc, na.rm = TRUE),
    aqi_daily = mean(aqi, na.rm = TRUE),
    .groups = "drop"
  )

# Monthly average
pm25_monthly <- pm25_daily %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Month) %>%
  summarize(
    raw_conc_monthly = mean(raw_conc_daily, na.rm = TRUE),
    aqi_monthly = mean(aqi_daily, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r, PM2.5 Trends Over Time}
ggplot(pm25_daily, aes(x = Date, y = raw_conc_daily)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Daily PM2.5 Concentrations in Ulaanbaatar (µg/m³)",
    x = "Date",
    y = "Daily Mean PM2.5"
  ) +
  theme_minimal()
```

```{Analysis, Concentration in Ulaanbaatar}
#The plot titled "Daily PM2.5 Concentrations in Ulaanbaatar (µg/m³)" shows a time series of daily mean PM2.5 levels from 2015 through early 2025. The data reveals a recurring pattern of sharp spikes in pollution levels each year, with the highest concentrations consistently occurring during the colder months. These peaks reach over 600 µg/m³ in some years, notably in 2016 and 2017, suggesting severe air quality events. Although the magnitude of these spikes varies year to year, the pattern of elevated concentrations in specific periods remains consistent throughout the decade, indicating persistent seasonal pollution events.


```

```{r, Missing Data Patterns in PM2.5}

#We evaluate missingness by calculating how many months each year had incomplete PM2.5 daily records.

pm25_yearly_missing <- pm25_monthly %>%
  mutate(Year = year(Month)) %>%
  group_by(Year) %>%
  summarize(
    months_with_missing = sum(is.na(raw_conc_monthly)),
    .groups = "drop"
  )

ggplot(pm25_yearly_missing, aes(x = Year, y = months_with_missing)) +
  geom_col(fill = "tomato") +
  labs(
    title = "Figure 2. Number of Months with Missing PM2.5 Data",
    x = "Year",
    y = "Months with Missing PM2.5"
  ) +
  theme_minimal()
```

```{Analysis, Number of Months with Missing PM2.5 Data}
#Our analysis in Figure 2 reveals that data gaps were mostly depicted in 2016, with 10 months lacking complete PM2.5 measurements. Additionally, one month in 2017 also shows missing data. From 2018 onward, there have been no recorded months with missing data. Therefore, the team had to analyse months with missing PM2.5 data
```

```{r, Distribution of Monthly PM2.5 Concentrations}

#We also explore the variability of monthly PM2.5 to check for seasonal peaks and outliers.

ggplot(pm25_monthly, aes(y = raw_conc_monthly)) +
  geom_boxplot(outlier.color = "red") +
  labs(
    title = "Figure 3. Distribution of Monthly PM2.5 Concentrations",
    y = "Monthly Mean PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```
```{Analysis, Distribution of Monthly PM2.5 Concentrations}
#The boxplot summarizes the variation in monthly mean PM2.5 levels in Ulaanbaatar. The plot illustrates that the median monthly PM2.5 concentration is just above 50 µg/m³, with the interquartile range (IQR)—representing the middle 50% of values—extending from approximately 20 to 110 µg/m³. Several outliers are displayed above the upper whisker, with some months reporting concentrations exceeding 300 µg/m³. These outliers likely indicate extreme pollution events occurring in particular months, aligning with the sharp peaks in Figure 1. The extended upper whisker signifies a right-skewed distribution, implying that while high PM2.5 values are less common, they can be significantly elevated when they do occur. In summary, this boxplot demonstrates that most monthly averages remain substantially above the WHO 24-hour guideline of 15 µg/m³, with a few months experiencing extremely high levels, underscoring the seriousness and fluctuations of air pollution in Ulaanbaatar.
```


```{r, Birth Outcomes Over Time}

#Before merging exposure and outcomes, we visualize trends in live births and low birth weight births separately.

# Plot live births over time
ggplot(births_merged, aes(x = Date, y = Live_Births)) +
  geom_line(color = "darkgreen") +
  labs(
    title = "Figure 4. Monthly Live Births in Ulaanbaatar",
    x = "Date",
    y = "Number of Live Births"
  ) +
  theme_minimal()

# Plot low birth weight births over time
ggplot(births_merged, aes(x = Date, y = Low_Birth_Weight)) +
  geom_line(color = "purple") +
  labs(
    title = "Figure 5. Monthly Low Birth Weight Births in Ulaanbaatar",
    x = "Date",
    y = "Number of LBW Births"
  ) +
  theme_minimal()
```
```{Analysis, Monthly Low Birth Weight Births in Ulaanbaatar}
#This chart shows that between 2016 and approximately 2021, the rate of LBW births was notably high and fluctuated significantly, consistently surpassing 160 births per month, occasionally exceeding 200. However, starting in 2022, a clear downward trend is evident, with both the average rate of LBW births and their levels of variability decreasing. By 2024, the number of LBW births often dips below 150 per month, marking some of the lowest figures recorded during the observed timeframe.

```
#Merge Exposure and Outcome Data
We link monthly PM2.5 exposure data with corresponding monthly birth outcomes. Therefore, in this section, we merge exposure and outcome datasets, calculate the low birth weight (LBW) rate, and begin preliminary modeling to investigate associations between PM2.5 exposure and birth outcomes.

```{r}
# Merge datasets by Date
full_data <- births_merged %>%
  left_join(pm25_monthly, by = c("Date" = "Month")) %>%
  arrange(Date)
```

#Calculate Low Birth Weight Rate
# here we create a new variable:

LBW_rate = (Low Birth Weight / Live Births) * 100

```{r}

# Calculate LBW rate as percentage
full_data <- full_data %>%
  mutate(LBW_rate = 100 * Low_Birth_Weight / Live_Births)

```

#Summary Statistics

```{r}

summary_stats <- full_data %>%
  summarise(
    Mean_LBW_Rate = mean(LBW_rate, na.rm = TRUE),
    Median_LBW_Rate = median(LBW_rate, na.rm = TRUE),
    Mean_PM25 = mean(raw_conc_monthly, na.rm = TRUE),
    Median_PM25 = median(raw_conc_monthly, na.rm = TRUE)
  )

summary_stats %>%
  kable(caption = "Table 1. Summary Statistics of LBW Rate and PM2.5", digits = 2) %>%
  kable_styling(full_width = FALSE)

```

#Scatterplot: PM2.5 vs LBW Rate
# Now we visualize the bivariate relationship before modeling.

```{r}
ggplot(full_data, aes(x = raw_conc_monthly, y = LBW_rate)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Figure 6. Relationship between Monthly PM2.5 and LBW Rate",
    x = "Monthly PM2.5 (µg/m³)",
    y = "Low Birth Weight Rate (%)"
  ) +
  theme_minimal()
```
```{Analysis, Relationship between Monthly PM2.5 and LBW Rate}
#The plot reveals a positive trend, indicating that higher levels of PM2.5 are associated with slightly higher rates of low birth weight. Although the slope of the regression line is not steep, the upward direction suggests that increases in air pollution may contribute to adverse birth outcomes. The shaded area around the line represents the 95% confidence interval, showing some uncertainty in the estimate, but the trend remains visible.
```

#Simple Linear Regression
#We fit an initial simple model:
LBW Rate ~ Monthly PM2.5 Concentration

```{r}
model_simple <- lm(LBW_rate ~ raw_conc_monthly, data = full_data)

tidy(model_simple) %>%
  kable(caption = "Table 2. Simple Linear Regression of LBW Rate on PM2.5", digits = 3) %>%
  kable_styling(full_width = FALSE)
```
```{Analysis,Simple Linear Regression of LBW Rate on PM2.5}
#```{Analysis, Relationship between Monthly PM2.5 and LBW Rate}
#The plot shows a positive trend, indicating that higher levels of PM2.5 are associated with slightly higher rates of low birth weight. Although the slope of the regression line is not steep, the upward direction suggests that increases in air pollution may contribute to adverse birth outcomes. The shaded area around the line represents the 95% confidence interval, showing some uncertainty in the estimate, but the trend remains visible.
```
```
#Lagged Exposure Variables
We construct lagged PM2.5 variables to capture exposures during previous months relative to the birth month.

```{r}

full_data <- full_data %>%
  arrange(Date) %>%
  mutate(
    pm25_lag0 = raw_conc_monthly,            # Current month
    pm25_lag1 = lag(raw_conc_monthly, 1),     # 1 month before
    pm25_lag2 = lag(raw_conc_monthly, 2),     # 2 months before
    pm25_lag3 = lag(raw_conc_monthly, 3)      # 3 months before
  )
  
```


```{r, Distributed Lag Model}
# Now We fit a distributed lag linear model:
#LBW Rate ~ PM2.5 exposure in current and previous 3 months

model_lag <- lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3, data = full_data)

tidy(model_lag) %>%
  kable(caption = "Table 3. Distributed Lag Model of LBW Rate on PM2.5 Exposure", digits = 3) %>%
  kable_styling(full_width = FALSE)
```

```{Analysis of distributed lag}
#The distributed lag model does not find statistically significant evidence that monthly PM2.5 exposure from 0 to 3 months prior has a measurable impact on LBW rate in this dataset. However, the near-significant result for lag 3 could indicate a potential delayed effect worth further investigation, perhaps with more data or a different model specification.
```

```{r}
#Cumulative Exposure Variables: We also create cumulative exposure averages to smooth over lag periods, as fetal development may be affected by sustained pollution rather than isolated monthly spikes.

full_data <- full_data %>%
  mutate(
    pm25_cum12 = (pm25_lag1 + pm25_lag2) / 2,          # 1–2 months average
    pm25_cum123 = (pm25_lag1 + pm25_lag2 + pm25_lag3) / 3  # 1–3 months average
  )
  
#Analysis: Here, we also get the same result we got for the distributed lag model
```


```{r, Regression on Cumulative Exposure}
#Finally, we fit simple models of LBW Rate on cumulative exposures:

model_cum12 <- lm(LBW_rate ~ pm25_cum12, data = full_data)
model_cum123 <- lm(LBW_rate ~ pm25_cum123, data = full_data)

bind_rows(
  tidy(model_cum12) %>% mutate(Model = "Cumulative Lag 1-2"),
  tidy(model_cum123) %>% mutate(Model = "Cumulative Lag 1-3")
) %>%
  select(Model, term, estimate, std.error, statistic, p.value) %>%
  kable(caption = "Table 4. Regression of LBW Rate on Cumulative PM2.5 Exposures", digits = 3) %>%
  kable_styling(full_width = FALSE)

#Analysis:Both cumulative models show a very small estimated increase in LBW rate (0.001 percentage points) per unit increase in cumulative PM2.5 exposure, but neither result is statistically significant. This suggests that, based on the available data, cumulative PM2.5 exposure in the last 2–3 months of pregnancy is not strongly or consistently associated with changes in LBW rate.
```

 
```{r,Multicollinearity Diagnostics}

#Now we assess multicollinearity among the lagged PM2.5 variables using VIFs.A VIF greater than 5–10 indicates problematic multicollinearity.

# Load car package if not already
library(car)

# Fit lagged model again if needed
model_lag <- lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3, data = full_data)

# Calculate VIFs
vif_values <- vif(model_lag)

# Display neatly
vif_values %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  rename(VIF = ".") %>%
  kable(caption = "Table 5. Variance Inflation Factors for Lagged PM2.5 Model", digits = 2) %>%
  kable_styling(full_width = FALSE)

#Analysis:The high VIF values (especially for lags 1 and 2) suggest that the lagged PM2.5 variables are strongly correlated with one another, which can inflate standard errors and make it difficult to detect statistically significant effects in your model.

```

**Methodology and Limitations**

#Statistical Approach and Results
**The analysis began with a simple linear regression to test the association between monthly PM2.5 concentrations and LBW rates in Ulaanbaatar. This initial model, which regressed LBW rates on same-month PM2.5 levels, produced a weak and statistically non-significant relationship [p = 0.113].**


**To account for the possibility that air pollution during pregnancy may affect birth outcomes with a delay, a distributed lag model was applied, incorporating PM2.5 exposures lagged by 0–3 months. However, this approach revealed high multicollinearity among lagged variables (variance inflation factors greater than 8 for some lags), which inflated standard errors and made it difficult to isolate the effect of any single month’s exposure.**


**Cumulative exposure models were then constructed by averaging PM2.5 over biologically plausible windows (e.g., 1–2 months and 1–3 months before birth) to capture sustained exposure during critical gestational periods. Despite this, the results remained non-significant, with small coefficients beta = 0.001, high p values [p = 0.158 for 1-2 months, p = 0.390 for 1-3 months]**


**These findings contrast with robust associations reported in the literature, where studies using individual-level data and precise gestational timing (such as aligning PM2.5 exposure with the second or third trimester) consistently found significant effects (Amnuaylojaroen & Saokaew, 2024; Zhang et al., 2019). For example, multinational meta-analyses and cohort studies that adjusted for confounders like socioeconomic status and healthcare access have reported a 5–8% increase in LBW risk per 10 µg/m³ rise in PM2.5 (Lee & Holm, 2022). The lack of significance in the current analysis likely stems from methodological limitations, including the use of aggregated monthly data (which cannot pinpoint critical gestational windows), unmeasured confounders (such as maternal age and poverty), and ecological bias.**

**Discussion**
#The present analysis did not identify a statistically significant relationship between monthly PM2.5 concentrations and LBW rates in Ulaanbaatar, regardless of whether exposure was considered in the same month, in previous months, or as a cumulative average. This outcome contrasts with findings from other regions, where higher air pollution during pregnancy has been linked to increased risk of LBW, especially in low-income or high-exposure settings (Amnuaylojaroen & Saokaew, 2024; Lee & Holm, 2022; Zhang et al., 2019). The discrepancy likely reflects differences in data structure and methodology. Specifically, the use of monthly aggregated data in this study limited the ability to align pollution exposure precisely with the most sensitive periods of pregnancy. Additionally, the absence of individual-level data and adjustment for important confounders, such as socioeconomic status and access to healthcare, may have masked subtle or time-specific effects of PM2.5 exposure.

#Limitations
Several limitations should be considered. First, monthly aggregated data limited the ability to pinpoint exposure during the most critical weeks of pregnancy, which is essential in air pollution and birth outcome research (Amnuaylojaroen & Saokaew, 2024; Zhang et al., 2019). Second, important confounders such as maternal age, income, and access to healthcare were not included, even though these factors are known to influence both pollution exposure and birth outcomes (Lee & Holm, 2022). Third, reliance on group-level data rather than individual pregnancy records may have introduced ecological bias. Fourth, high correlation between PM2.5 levels in adjacent months made it difficult to determine which period of exposure had the strongest effect. Finally, the sample size and time span may not have been sufficient to detect small or modest effects. Future research should use data that allow for more accurate timing of exposure and include additional variables to better understand how air pollution affects birth outcomes in Mongolia.


**Key findings**
- PM2.5 concentrations in Ulaanbaatar showed strong seasonal variation, with extreme peaks in winter months.
- The highest pollution levels occurred in 2016 and 2017, exceeding 600 µg/m³ on some days.
- Monthly low birth weight (LBW) rates showed a declining trend starting in 2022.
- Simple and lagged linear models did not detect a statistically significant association between PM2.5 exposure and LBW rate.
- Multicollinearity among lagged exposure variables was high, limiting the interpretability of individual lag effects.


**conclusion**
While elevated PM2.5 exposure in Ulaanbaatar is clearly a recurring and serious public health concern, this analysis does not find strong statistical evidence that short-term monthly or lagged PM2.5 exposure is independently associated with changes in LBW rates. Future studies should explore longer exposure windows and consider individual-level birth data to strengthen causal inference as we lacked the birth data here.

**References**
Amnuaylojaroen, T., & Saokaew, S. (2024). Prenatal PM2.5 exposure and its association with low birth weight: A systematic review and meta-analysis. Toxics, 12, 446.
Lee, J. R., & Holm, S. M. (2022). The association between ambient PM2.5 and low birth weight in California. International Journal of Environmental Research and Public Health, 19, 13554.
Zhang, Y., Wang, J., Chen, L., et al. (2019). Ambient PM2.5 and clinically recognized early pregnancy loss: A case-control study with spatiotemporal exposure predictions. Environment International, 126, 422–429.






