---
title: "Impact of PM2.5 Exposure on Low Birth Weight in Ulaanbaatar (2016-2025)"
author: "Ahmedi, Barua, Bhuyan, Karayel"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default

---
```{r}
knitr::opts_knit$set(root.dir = here::here())
```



```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(readr)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(kableExtra)
library(scales)
library(broom)
library(ggpubr)
library(here)
library(janitor)
library(forecast)
```

# 1. Load and Clean Data

```{r read-data}
# Read birth weight and live births data
birth_weight_low <- read.csv(here("Data/Raw/BIRTH WEIGTH LOWER THAN 2500 GRAMS.csv"), stringsAsFactors = TRUE)

live_births <- read.csv(here("./Data/Raw/LIVE BIRTHS.csv"), stringsAsFactors = TRUE)

# Clean live births: remove commas (if any) and converting to numeric()
# Might not need this as a visula check

live_births_clean <- live_births
 
 for (col in names(live_births_clean)[-1]) {
   live_births_clean[[col]] <- as.numeric(gsub(",", "", live_births_clean[[col]]))
 }

# The data is wide, need to convert the data to long format
birth_weight_low_long <- birth_weight_low %>% 
  pivot_longer(-Aimag, 
               names_to = "Month", 
               values_to = "Low_Birth_Weight")

live_births_long <- live_births_clean %>% 
  pivot_longer(-Aimag, 
               names_to = "Month", 
               values_to = "Live_Births")

# Merge two datasets
births_merged <- left_join(birth_weight_low_long,
                           live_births_long, 
                           by = c("Aimag", "Month"))
# Removing "X" from month names
birth_weight_low_long <- birth_weight_low_long %>%
  mutate(Month = gsub("^X", "", Month))

live_births_long <- live_births_long %>%
  mutate(Month = gsub("^X", "", Month))

# Merging two datasets
births_merged <- left_join(birth_weight_low_long, live_births_long, by = c("Aimag", "Month"))

# Creating Date column
births_merged <- births_merged %>%
  mutate(Date = ym(Month)) %>%
  select(Aimag, Date, Low_Birth_Weight, Live_Births)

# # Quick checks
# str(births_merged)
# colSums(is.na(births_merged))
# summary(births_merged)
# class(births_merged)

```



```{r read-pm25}

# 1. Read and Combine All PM2.5 Files
years <- 2015:2025
pm25_files <- paste0("./Data/Raw/Ulaanbaatar_PM2.5_", years, "_YTD.csv")
names(pm25_files) <- years

# Read and bind all
pm25_all <- map_dfr(pm25_files, ~ read_csv(.x, show_col_types = FALSE))

# 2. Convert all -999 to NA across numeric columns only
pm25_all <- pm25_all %>%
  mutate(across(where(is.numeric), ~ na_if(., -999)))


# Now I need to convert all the hourly, daily, monthly and yearly data into a DateTime object. The plan is to separate hourly data and take average of hours to make it daily and same to make it monthly. 

pm25_all <- pm25_all %>%
  clean_names() # Date(LT) was giving all troubles. used janitor package to rename to clean                      variable name. it cleaned all the variables name

pm25_all <- pm25_all %>%
  rename(DateTime = date_lt) %>%  
  mutate(
    DateTime = parse_date_time(DateTime, orders = "ymd IMp"),
    Date     = date(DateTime)
  )

# Now I will create 3 dataset, hourly, daily and monthly just to make sure if anything goes wrong I can come back
pm25_hourly <- pm25_all

# DAILY aggregation
# Did not use nowcast as it is smoothed data. used raw concentration

# DAILY aggregation
pm25_daily <- pm25_hourly %>%
  mutate(
    Date     = date(DateTime)                # extract YYYY-MM-DD
  ) %>%
  group_by(Date) %>%
  summarize(
    raw_conc_daily    = mean(raw_conc, na.rm = TRUE),
    aqi_daily         = mean(aqi,      na.rm = TRUE),
    hours_reported    = n(),
    hours_missing_raw = sum(is.na(raw_conc)),
    hours_missing_aqi = sum(is.na(aqi)),
    .groups = "drop"
  ) %>%
  mutate(
    DateTime = as_datetime(Date)             # midnight timestamps
  )

# MONTHLY aggregation
pm25_monthly <- pm25_daily %>%
  mutate(
    Month = floor_date(Date, "month")        # first day of each month
  ) %>%
  group_by(Month) %>%
  summarize(
    raw_conc_monthly   = mean(raw_conc_daily, na.rm = TRUE),
    aqi_monthly        = mean(aqi_daily,      na.rm = TRUE),
    days_reported      = n(),
    days_missing_raw   = sum(is.na(raw_conc_daily)),
    days_missing_aqi   = sum(is.na(aqi_daily)),
    .groups = "drop"
  ) %>%
  mutate(
    DateTime = as_datetime(Month)            # first‐of‐month timestamps
  )

# YEARLY aggregation
pm25_yearly <- pm25_monthly %>%
  mutate(
    Year = year(Month)
  ) %>%
  group_by(Year) %>%
  summarize(
    raw_conc_yearly    = mean(raw_conc_monthly, na.rm = TRUE),
    aqi_yearly         = mean(aqi_monthly,      na.rm = TRUE),
    months_reported    = n(),
    months_missing_raw = sum(days_missing_raw > 0),
    months_missing_aqi = sum(days_missing_aqi > 0),
    .groups = "drop"
  ) %>%
  mutate(
    DateTime = ymd(paste0(Year, "-01-01"))    # Jan 1 of each year
  )


# Visuallizing pattern

ggplot(pm25_daily, aes(x = Date, y = raw_conc_daily)) +
  geom_line() +
  labs(
    title = "Daily PM 2.5 Concentrations (µg/m3)",
    x     = "Date",
    y     = "Daily mean raw_conc"
  ) +
  theme_minimal()

# Treating missing value: 
# visualizing missing value

# Bar chart: number of months with ≥1 missing day each year
# Create the pm25_yearly_missing summary
pm25_yearly_missing <- pm25_monthly %>%
  mutate(Year = year(Month)) %>%
  group_by(Year) %>%
  summarize(
    total_months               = n(),
    months_with_missing_days   = sum(days_missing_raw > 0),
    total_missing_days         = sum(days_missing_raw),
    .groups = "drop"
  )

# Plot
ggplot(pm25_yearly_missing, aes(x = Year, y = months_with_missing_days)) +
  geom_col(fill = "tomato") +
  labs(
    title = "Number of Months with Missing PM2.5 Data by Year",
    x     = "Year",
    y     = "Months with ≥1 Missing Day"
  ) +
  theme_minimal()

# It looks like there is 9 month of missing data in 2015. We need to consider it afterwards.

# Boxplot of monthly series to spot outliers
ggplot(pm25_monthly, aes(y = raw_conc_monthly)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1) +
  labs(
    title = "Distribution of Monthly PM2.5",
    y     = "Monthly mean PM2.5 (µg/m3)"
  ) +
  theme_minimal()

# Looks like does not have a lot of outliers. we can ignore


#write_csv(pm25_all, here("Data", "Processed", "pm25_all.csv"))



# Final NA check
summary(pm25_monthly)
colSums(is.na(pm25_monthly))




```

```{r merge-data}
# Merge PM2.5 with births
full_data <- births_merged %>%
  left_join(
    pm25_monthly,
    by = c("Date" = "Month")
  ) %>%
  arrange(Date)


full_data %>% 
  select(Date, Aimag, Low_Birth_Weight, Live_Births, raw_conc_monthly, aqi_monthly)
```

```{r Sumamry Staistics}


# Summary for birth outcomes
births_summary <- full_data %>%
  summarise(
    Mean_LBW       = mean(Low_Birth_Weight, na.rm = TRUE),
    Median_LBW     = median(Low_Birth_Weight, na.rm = TRUE),
    Min_LBW        = min(Low_Birth_Weight, na.rm = TRUE),
    Max_LBW        = max(Low_Birth_Weight, na.rm = TRUE),
    SD_LBW         = sd(Low_Birth_Weight, na.rm = TRUE),
    N_LBW          = sum(!is.na(Low_Birth_Weight)),
    
    Mean_Live      = mean(Live_Births, na.rm = TRUE),
    Median_Live    = median(Live_Births, na.rm = TRUE),
    Min_Live       = min(Live_Births, na.rm = TRUE),
    Max_Live       = max(Live_Births, na.rm = TRUE),
    SD_Live        = sd(Live_Births, na.rm = TRUE),
    N_Live         = sum(!is.na(Live_Births))
  )

births_summary %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("Statistic") %>%
  rename(Value = V1) %>%
  kable(caption = "Summary of Birth Outcomes", digits = 2) %>%
  kable_styling(full_width = FALSE)

# 2. Summary for PM2.5 exposure
pm25_summary <- full_data %>%
  summarise(
    Mean_PM25     = mean(raw_conc_monthly, na.rm = TRUE),
    Median_PM25   = median(raw_conc_monthly, na.rm = TRUE),
    Min_PM25      = min(raw_conc_monthly, na.rm = TRUE),
    Max_PM25      = max(raw_conc_monthly, na.rm = TRUE),
    SD_PM25       = sd(raw_conc_monthly, na.rm = TRUE),
    N_PM25        = sum(!is.na(raw_conc_monthly)),

    Mean_AQI      = mean(aqi_monthly, na.rm = TRUE),
    Median_AQI    = median(aqi_monthly, na.rm = TRUE),
    Min_AQI       = min(aqi_monthly, na.rm = TRUE),
    Max_AQI       = max(aqi_monthly, na.rm = TRUE),
    SD_AQI        = sd(aqi_monthly, na.rm = TRUE),
    N_AQI         = sum(!is.na(aqi_monthly))
  )

pm25_summary %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("Statistic") %>%
  rename(Value = V1) %>%
  kable(caption = "Summary of Monthly PM2.5 Exposure", digits = 2) %>%
  kable_styling(full_width = FALSE)


```


# Descriptive Statistics
```{r}

# Compute low birth weight rate (%)
full_data <- full_data %>%
  mutate(
    LBW_rate = 100 * Low_Birth_Weight / Live_Births
  )

# 2. Summary table of exposure and outcome
summary_tbl <- full_data %>%
  summarise(
    Mean_PM25     = mean(raw_conc_monthly, na.rm = TRUE),
    SD_PM25       = sd(raw_conc_monthly, na.rm = TRUE),
    Mean_LBWrate  = mean(LBW_rate, na.rm = TRUE),
    SD_LBWrate    = sd(LBW_rate, na.rm = TRUE),
    N             = n()
  ) %>%
  pivot_longer(everything(), names_to="Metric", values_to="Value")

summary_tbl %>%
  kable(caption="Summary of PM2.5 and LBW Rate", digits=2) %>%
  kable_styling(full_width=FALSE)

# Scatter + trend line
ggplot(full_data, aes(x = raw_conc_monthly, y = LBW_rate)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE, color="blue") +
  labs(
    title = "Low Birth Weight Rate vs. Monthly PM2.5",
    x     = "PM2.5 (µg/m³)",
    y     = "LBW Rate (%)"
  ) +
  theme_minimal()
```
#  Linear regression
```{r}
model <- lm(LBW_rate ~ raw_conc_monthly, data = full_data)
tidy(model) %>%
  kable(caption="Regression of LBW Rate on PM2.5", digits=3) %>%
  kable_styling(full_width=FALSE)
# Create proportion of low birth weight
full_data <- full_data %>%
  mutate(Proportion_LBW = Low_Birth_Weight / Live_Births)
```


Exposure timing plays a key role in accurately identifying how air pollution affects birth outcomes. Our initial analysis showed a weak and non-significant association between monthly PM2.5 levels and low birth weight rates (p = 0.113). This limited result probably occurred because measuring air pollution in the same month as birth ignores the critical periods during pregnancy when exposure actually affects fetal growth.

To address this issue, we will use lagged regression methods. Lagged regression aligns PM2.5 exposure data with biologically relevant periods earlier in pregnancy, specifically during critical developmental windows like the second and third trimesters. Aligning exposure measurements with these sensitive gestational periods, similar to optimizing exposure timing in optical measurement systems (Wang et al., 2020; Li et al., 2021; Zhang et al., 2019), helps to minimize measurement uncertainty, reduce confounding from seasonal variations, and accurately identify the true health impacts of PM2.5 exposure on birth weight outcomes.

References:
Li et al., 2021; Wang et al., 2020; Zhang et al., 2019.

```{r}

# making sure it's sorted by date
full_data <- full_data %>% 
  arrange(Date)

# creating lags 
full_data <- full_data %>% 
  mutate(
    pm25_lag0 = raw_conc_monthly,
    pm25_lag1 = lag(raw_conc_monthly, 1),
    pm25_lag2 = lag(raw_conc_monthly, 2),
    pm25_lag3 = lag(raw_conc_monthly, 3)
  )

# fitting a “distributed‐lag” linear model on the LBW rate
model_lag <- lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3,
                data = full_data)

# inspecting the month‐specific coefficients
tidy(model_lag) %>% 
  kable(digits=3, caption="Lagged PM₂.₅ Effects on LBW Rate") %>%
  kable_styling(full_width=FALSE)

# # Check correlations
cor(full_data %>% select(pm25_lag0:pm25_lag3), use = "pairwise.complete.obs")

# Or VIFs
library(car)
vif(lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3,
       data = full_data))



# Cumulative two‐month average
full_data <- full_data %>%
  mutate(pm25_cum23 = (pm25_lag2 + pm25_lag3) / 2)

lm(LBW_rate ~ pm25_cum23, data = full_data) %>% broom::tidy()

dlm <- lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3,
          data = full_data)
linearHypothesis(dlm, c("pm25_lag0 = 0",
                        "pm25_lag1 = 0",
                        "pm25_lag2 = 0",
                        "pm25_lag3 = 0"))

## Collapse to 2- and 3-month cumulative PM₂.₅ averages and re-fit

library(dplyr)
library(broom)
library(knitr)
library(kableExtra)

# 1) Create cumulative exposure variables
full_data <- full_data %>%
  arrange(Date) %>%
  mutate(
    pm25_cum12  = (pm25_lag1 + pm25_lag2) / 2,          # average of 1- and 2-month lags
    pm25_cum123 = (pm25_lag1 + pm25_lag2 + pm25_lag3) / 3  # average of 1-, 2-, 3-month lags
  )

# 2) Fit separate linear models
model_cum12  <- lm(LBW_rate ~ pm25_cum12,  data = full_data)
model_cum123 <- lm(LBW_rate ~ pm25_cum123, data = full_data)

# 3) Summarize and display results side by side
results <- bind_rows(
  tidy(model_cum12)  %>% mutate(model = "cum12"),
  tidy(model_cum123) %>% mutate(model = "cum123")
) %>%
  select(model, term, estimate, std.error, statistic, p.value)

results %>%
  kable(
    caption = "Regression of LBW Rate on Cumulative PM2.₅ Averages",
    digits  = 3
  ) %>%
  kable_styling(full_width = FALSE)





```


---

# End of Analysis

