---
title: "Impact of PM2.5 Exposure on Low Birth Weight in Ulaanbaatar from 2016–2025"
author: "Ahmedi, Barua, Bhuyan, Karayel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
library(lubridate)
library(readr)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(kableExtra)
library(scales)
library(broom)
library(ggpubr)
library(here)
library(janitor)
library(forecast)
library(car)
```
# Rationale and Research Questions

Air pollution, particularly PM2.5, remains a significant public health concern in Ulaanbaatar.  
This project examines whether exposure to elevated PM2.5 concentrations during pregnancy is associated with an increased risk of low birth weight (LBW) among newborns between 2016 and 2025.

**Research Questions:**
- What are the seasonal and long-term trends in PM2.5 concentrations in Ulaanbaatar?
- Is there a measurable association between PM2.5 exposure levels and low birth weight rates?
- Does the timing of PM2.5 exposure during pregnancy (examined through lagged exposure models) influence birth outcomes?

---

# Dataset Information

## Data Sources

- **PM2.5 Data**: Hourly PM2.5 concentration measurements collected from Ulaanbaatar monitoring stations from 2015 to 2025.
- **Birth Outcomes Data**: Monthly records of total live births and low birth weight births for the same period.

## Data Cleaning and Wrangling

- **PM2.5 data**:
  - Combined annual CSV files across multiple years.
  - Replaced invalid `-999` entries with `NA`.
  - Aggregated hourly readings into daily and monthly averages to align with birth record reporting.

- **Birth outcomes data**:
  - Reshaped from wide format to long format for time series analysis.
  - Merged live births and low birth weight counts into a unified dataset.

- **Final merged dataset**:
  - Monthly PM2.5 exposure data was linked with corresponding monthly birth outcome data by date.

## Dataset Structure

| Dataset           | Key Variables                         | Notes                            |
|-------------------|----------------------------------------|----------------------------------|
| PM2.5 Pollution    | DateTime, Raw Concentration, AQI       | Aggregated to monthly averages   |
| Birth Outcomes     | Date, Aimag, Low Birth Weight count, Live Births | Cleaned and merged records |


---

# Data Loading

We now load and prepare the raw PM2.5 and birth outcome datasets as described previously.

```{r load-data}
# Define file paths for PM2.5 data
years <- 2015:2025
pm25_files <- paste0("Data/Raw/Ulaanbaatar_PM2.5_", years, "_YTD.csv")
names(pm25_files) <- years

# Read and bind all PM2.5 files
pm25_all <- map_dfr(pm25_files, ~ read_csv(.x, show_col_types = FALSE))

# Clean and convert -999 to NA
pm25_all <- pm25_all %>%
  mutate(across(where(is.numeric), ~ na_if(., -999))) %>%
  clean_names() %>%
  rename(DateTime = date_lt) %>%
  mutate(
    DateTime = parse_date_time(DateTime, orders = "ymd IMp"),
    Date     = as_date(DateTime)
  )

# Load birth outcome data
birth_weight_low <- read_csv("Data/Raw/BIRTH WEIGTH LOWER THAN 2500 GRAMS.csv")
live_births      <- read_csv("Data/Raw/LIVE BIRTHS.csv")

# Clean numeric columns in live births (in case of commas)
for (col in names(live_births)[-1]) {
  live_births[[col]] <- as.numeric(gsub(",", "", live_births[[col]]))
}

# Reshape both datasets to long format
birth_weight_long <- birth_weight_low %>%
  pivot_longer(-Aimag, names_to = "Month", values_to = "Low_Birth_Weight") %>%
  mutate(Month = gsub("^X", "", Month))

live_births_long <- live_births %>%
  pivot_longer(-Aimag, names_to = "Month", values_to = "Live_Births") %>%
  mutate(Month = gsub("^X", "", Month))

# Merge and finalize
births_merged <- left_join(birth_weight_long, live_births_long, by = c("Aimag", "Month")) %>%
  mutate(Date = ym(Month)) %>%
  select(Aimag, Date, Low_Birth_Weight, Live_Births)

```

# Exploratory Analysis
We begin by examining the structure, quality, and temporal characteristics of the air pollution data.

##Daily and Monthly Aggregation of PM2.5

```{r}
# Daily average
pm25_daily <- pm25_all %>%
  group_by(Date) %>%
  summarize(
    raw_conc_daily = mean(raw_conc, na.rm = TRUE),
    aqi_daily = mean(aqi, na.rm = TRUE),
    .groups = "drop"
  )

# Monthly average
pm25_monthly <- pm25_daily %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Month) %>%
  summarize(
    raw_conc_monthly = mean(raw_conc_daily, na.rm = TRUE),
    aqi_monthly = mean(aqi_daily, na.rm = TRUE),
    .groups = "drop"
  )
```


#PM2.5 Trends Over Time

```{r}
ggplot(pm25_daily, aes(x = Date, y = raw_conc_daily)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Daily PM2.5 Concentrations in Ulaanbaatar (µg/m³)",
    x = "Date",
    y = "Daily Mean PM2.5"
  ) +
  theme_minimal()
```

#Missing Data Patterns in PM2.5
We evaluate missingness by calculating how many months each year had incomplete PM2.5 daily records.

```{r}

pm25_yearly_missing <- pm25_monthly %>%
  mutate(Year = year(Month)) %>%
  group_by(Year) %>%
  summarize(
    months_with_missing = sum(is.na(raw_conc_monthly)),
    .groups = "drop"
  )

ggplot(pm25_yearly_missing, aes(x = Year, y = months_with_missing)) +
  geom_col(fill = "tomato") +
  labs(
    title = "Figure 2. Number of Months with Missing PM2.5 Data",
    x = "Year",
    y = "Months with Missing PM2.5"
  ) +
  theme_minimal()
```

#Distribution of Monthly PM2.5 Concentrations
We also explore the variability of monthly PM2.5 to check for seasonal peaks and outliers.

```{r}
ggplot(pm25_monthly, aes(y = raw_conc_monthly)) +
  geom_boxplot(outlier.color = "red") +
  labs(
    title = "Figure 3. Distribution of Monthly PM2.5 Concentrations",
    y = "Monthly Mean PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```
#Birth Outcomes Over Time
Before merging exposure and outcomes, we visualize trends in live births and low birth weight births separately.


```{r}
# Plot live births over time
ggplot(births_merged, aes(x = Date, y = Live_Births)) +
  geom_line(color = "darkgreen") +
  labs(
    title = "Figure 4. Monthly Live Births in Ulaanbaatar",
    x = "Date",
    y = "Number of Live Births"
  ) +
  theme_minimal()

# Plot low birth weight births over time
ggplot(births_merged, aes(x = Date, y = Low_Birth_Weight)) +
  geom_line(color = "purple") +
  labs(
    title = "Figure 5. Monthly Low Birth Weight Births in Ulaanbaatar",
    x = "Date",
    y = "Number of LBW Births"
  ) +
  theme_minimal()
```

##Analysis 

In this section, we merge exposure and outcome datasets, calculate the low birth weight (LBW) rate, and begin preliminary modeling to investigate associations between PM2.5 exposure and birth outcomes.

#Merge Exposure and Outcome Data
We link monthly PM2.5 exposure data with corresponding monthly birth outcomes.

```{r}
# Merge datasets by Date
full_data <- births_merged %>%
  left_join(pm25_monthly, by = c("Date" = "Month")) %>%
  arrange(Date)
```


#Calculate Low Birth Weight Rate
We create a new variable:
LBW rate (%) = (Low Birth Weight / Live Births) * 100

```{r}

# Calculate LBW rate as percentage
full_data <- full_data %>%
  mutate(LBW_rate = 100 * Low_Birth_Weight / Live_Births)


```


#Summary Statistics

```{r}

summary_stats <- full_data %>%
  summarise(
    Mean_LBW_Rate = mean(LBW_rate, na.rm = TRUE),
    Median_LBW_Rate = median(LBW_rate, na.rm = TRUE),
    Mean_PM25 = mean(raw_conc_monthly, na.rm = TRUE),
    Median_PM25 = median(raw_conc_monthly, na.rm = TRUE)
  )

summary_stats %>%
  kable(caption = "Table 1. Summary Statistics of LBW Rate and PM2.5", digits = 2) %>%
  kable_styling(full_width = FALSE)

```

#Scatterplot: PM2.5 vs LBW Rate

We visualize the bivariate relationship before modeling.


```{r}
ggplot(full_data, aes(x = raw_conc_monthly, y = LBW_rate)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Figure 6. Relationship between Monthly PM2.5 and LBW Rate",
    x = "Monthly PM2.5 (µg/m³)",
    y = "Low Birth Weight Rate (%)"
  ) +
  theme_minimal()
```

#Simple Linear Regression
We fit an initial simple model:
LBW Rate ~ Monthly PM2.5 Concentration



```{r}

model_simple <- lm(LBW_rate ~ raw_conc_monthly, data = full_data)

tidy(model_simple) %>%
  kable(caption = "Table 2. Simple Linear Regression of LBW Rate on PM2.5", digits = 3) %>%
  kable_styling(full_width = FALSE)
```

#Lagged Exposure Variables
We construct lagged PM2.5 variables to capture exposures during previous months relative to the birth month.

```{r}

full_data <- full_data %>%
  arrange(Date) %>%
  mutate(
    pm25_lag0 = raw_conc_monthly,            # Current month
    pm25_lag1 = lag(raw_conc_monthly, 1),     # 1 month before
    pm25_lag2 = lag(raw_conc_monthly, 2),     # 2 months before
    pm25_lag3 = lag(raw_conc_monthly, 3)      # 3 months before
  )
```


#Distributed Lag Model
We fit a distributed lag linear model:
LBW Rate ~ PM2.5 exposure in current and previous 3 months


```{r}


model_lag <- lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3, data = full_data)

tidy(model_lag) %>%
  kable(caption = "Table 3. Distributed Lag Model of LBW Rate on PM2.5 Exposure", digits = 3) %>%
  kable_styling(full_width = FALSE)
```

#Cumulative Exposure Variables
We also create cumulative exposure averages to smooth over lag periods, as fetal development may be affected by sustained pollution rather than isolated monthly spikes.

```{r}
full_data <- full_data %>%
  mutate(
    pm25_cum12 = (pm25_lag1 + pm25_lag2) / 2,          # 1–2 months average
    pm25_cum123 = (pm25_lag1 + pm25_lag2 + pm25_lag3) / 3  # 1–3 months average
  )
```



#Regression on Cumulative Exposure

Finally, we fit simple models of LBW Rate on cumulative exposures:

```{r}
model_cum12 <- lm(LBW_rate ~ pm25_cum12, data = full_data)
model_cum123 <- lm(LBW_rate ~ pm25_cum123, data = full_data)

bind_rows(
  tidy(model_cum12) %>% mutate(Model = "Cumulative Lag 1-2"),
  tidy(model_cum123) %>% mutate(Model = "Cumulative Lag 1-3")
) %>%
  select(Model, term, estimate, std.error, statistic, p.value) %>%
  kable(caption = "Table 4. Regression of LBW Rate on Cumulative PM2.5 Exposures", digits = 3) %>%
  kable_styling(full_width = FALSE)

```

#Multicollinearity Diagnostics: Variance Inflation Factors (VIF)
We assess multicollinearity among the lagged PM2.5 variables using VIFs.
A VIF greater than 5–10 indicates problematic multicollinearity.


```{r}

# Load car package if not already
library(car)

# Fit lagged model again if needed
model_lag <- lm(LBW_rate ~ pm25_lag0 + pm25_lag1 + pm25_lag2 + pm25_lag3, data = full_data)

# Calculate VIFs
vif_values <- vif(model_lag)

# Display neatly
vif_values %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  rename(VIF = ".") %>%
  kable(caption = "Table 5. Variance Inflation Factors for Lagged PM2.5 Model", digits = 2) %>%
  kable_styling(full_width = FALSE)
```










